---
title: "Global ETF Map Paper"
output:
  html_notebook: default
---

In this presentation, we will construct a Shiny app that allows users to graph country ETF price data by clicking on a world map. It might be hard to visualize exactly what that means, so let's a have a quick look at the final app here: http://colorado.rstudio.com:3939/Global-ETF-Map/. 

From a substantive perspective, the app is simple in that it just displays price histories.  But our learnings  are diverse and broadly applicable: an introduction to the Notebooks-to-flexdashboards workflow and the reproducibility it engenders, spatial dataframe and map-making skills, adding our own stock data to a map, wiring up two HTML widgets to communicate with one another and creating a useful app for users who don't want to remember unintuitive ticker symbols.

How will we get there? We are going to use two Rmarkdown files. We start with a Notebook (which we are looking at right now) for data import, wrangling, and test visualizations. Next, we use a flexdashboard to build our interactive Shiny app. We are going to save objects from the Notebook into a .Rdat file, and then use that file in our app. The Notebook is fundamentally connected to our flexdashboard. 

Let’s get to the Notebook data importing/tidying/visualizing steps.

First, we will get ETF tickers, countries and year-to-date performance data into a dataframe. Note that the dataframe will not hold the price history data itself. Rather, it will hold simply the ticker symbols, country names and YTD percentages.

Next, we pass those ticker symbols to the getSymbols() function and download the ETF price histories. Advance warning: there are 42 country ETFs in this example and downloading 42 xts objects takes time and RAM.  It is not strictly necessary to pass all of those tickers to getSymbols() right now because the data will be downloaded on the fly when a user clicks on a country in our Shiny app. However, better to find the typos now than to have users discover an error in the app.

Once we have confirmed that our ticker symbols are valid, we add ETF data to the shapefile, making them accessible via clicks on the map. At this step, we will be thankful that when we created a dataframe in step 1, we used the same country names as appear on the map: that forethought will allow us to do an easy ‘merge’ of the data.

Once we have a shapefile with our ETF tickers added, we’ll save it to a .RDat file that we load into our Shiny app. That will conclude our time here in the Notebook and then we'll toggle over to the flexdashboard.

Without further adieu, the ishares ETF tickers, country names and year-to-date performance are pasted below.

```{r, message = FALSE}
# Let's build a dataframe to store these ticker symbols, country names and YTD numbers.
library(dplyr)

ticker <-  c("EWJ",  "EWZ",  "INDA", "FXI",  "EWG",  "EWC",  "EWY",  "EWT",  "EWU",  "EWH",  "EWA",
             "EWW",  "EWL",  "EWP", "EWS",  "EWI",  "EIDO", "ERUS", "ECH",  "EZA",  "THD",  "TUR",
             "EWD",  "EWQ",  "EWM",  "EPU",  "EWN",  "EPOL", "EPHE", "ENZL", "EIRL", "EWK",  "EIS",
             "EWO", "EDEN", "QAT", "UAE", "EFNL", "ENOR", "ICOL", "HEWY", "KSA")

name <-   c("Japan", "Brazil" ,"India", "China", "Germany" , "Canada", "Korea", "Taiwan", 
              "United Kingdom", "Hong Kong", "Australia", "Mexico", "Switzerland", "Spain", 
              "Singapore", "Italy", "Indonesia", "Russia", "Chile", "South Africa", "Thailand",  
              "Turkey", "Sweden", "France", "Malaysia", "Peru", "Netherlands", "Poland",
              "Philippines", "New Zealand", "Ireland", "Belgium", "Israel", "Austria","Denmark",
              "Qatar", "United Arab Emirates", "Finland", "Norway", "Colombia", "South Korea", 
              "Saudi Arabia")

ytd <- c(0.0358, 0.6314, -0.0140,  0.0721, -0.0289, 0.2198,  0.0729,  0.2029, -0.0467,  0.0897,
         0.0944, -0.1045, -0.0623, -0.0916,  0.0305, -0.1857,  0.1309,  0.3828,  0.1987,  0.1219,
         0.2458, -0.1053, -0.0052, -0.0199, -0.0410,  0.6015,  0.0017, -0.0481, -0.0408,  0.1394,
         -0.1183, -0.0428, -0.0432,  0.0462, -0.1078, -0.0244, 0.0570,  0.6397, -0.0146,  0.1424,
         0.1313,  0.0751) * 100

etf_ticker_country <- data_frame(ticker, name, ytd)

etf_ticker_country
```

 
The dataframe looks pretty good, though quite simple, and it’s fair to wonder why I bothered to highlight this step with it’s own code chunk. In fact, getting the clean ticker and country names was quite time-consuming, and that will often be the case: the most prosaic data import and tidying tasks can take a long time! Here is a good occasion to bring up reproducibility and work flow. Once you or your colleague has spent the time to get a clean dataframe with ticker and country names, we definitely want to make sure that no one else, including your future self, has to duplicate the effort for a future project. I put this step in it’s own code chunk so that the path back to the clean data would be as clear as possible. 

Now that we have the tickers in a dataframe column, we can use getSymbols() to import the price history of each ETF. We aren’t going to use the results of this import in the app. Rather, we are going to perform this import to test that we have the correct symbols, and that they play nicely with getSymbols().

```{r, warning = FALSE, message = FALSE, eval = FALSE}
# getSymbols is part of the 'quantmod' package.

library(quantmod)

# Using getSymbols to import the ETF price histories will take a minute or two or 
# five - 42 time series is a lot of data. 

getSymbols(etf_ticker_country$ticker, auto.assign = TRUE, warnings = FALSE)

# Let select just the closing prices of the ETFs and merge them into a list.
# We'll use lapply for that purpose. Again, this is for testing purposes. It's not 
# going into production in our app.

etf_prices <- do.call(merge, lapply(etf_ticker_country$ticker, function(x) Ad(get(x))))

#Change the column names to the country names from our dataframe above.

colnames(etf_prices) <- etf_ticker_country$name

# Take a peek at the last 5 rows of each of the time series, 
# just to make sure it looks complete.

tail(etf_prices, n = 5)
```


We have imported the adjusted prices of the country ETFs. Nothing too complicated here and again, our purpose was to test that the ticker symbols are correct. We are not going to be saving these prices for future use. 

Now it’s time to build a map of the Earth! 

First, we will need a shapefile that contains the spatial polygons for the countries of the world.  

The next code chunk will grab a shapefile from www.naturalearthdata.com.  That shapefile has the longitude and latitude coordinates for the world's countries and some data about them.

We'll then use the readOGR() function from the 'rgdal' package to load the shapefile into our global environment.


```{r, message = FALSE, warning = FALSE}
library(rgdal)
library(rmapshaper)
library(httr)

tmp <- tempfile()

httr::GET("http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/50m/cultural/ne_50m_admin_0_countries.zip", write_disk(tmp))

unzip(tmp, exdir = 'ne_50m_admin_0_countries')

world <- readOGR("./ne_50m_admin_0_countries", 'ne_50m_admin_0_countries', verbose = FALSE)
world <- ms_simplify(world)
```

Take a peek at the dataframe portion of the shapefile, and scroll to the right to see some interesting things like GDP estimates and economic development stages. It’s pretty nice that the shapefile contains some economic data for us. The other portion of the shapefile is the spatial data: longitude and latitude coordinates. If you’re not a cartographer, don’t worry about those.

```{r}
# For the sake of brevity, let's look at just the first 6 countries, 
# their GDP estimates and the stage of their economy.
head(world@data[c('name', 'gdp_md_est', 'economy')])
```


The ‘gdp_md_est’ column contains GDP estimates for each country. We’ll add some color to our map with shades of blue that are darker for higher GDPs and lighter for lower GDPs.

```{r, warning = FALSE}
library(leaflet)

# Create a palette with different shades of blue for different
# GDP estimates.

gdpPal <- colorQuantile("Blues", world$gdp_md_est, n = 20)
```

We want something to happen when a user clicks a country. How about a popup with country
name and stage of economic development? Again, that data is included in the shapefile we downloaded.

```{r}
# Make a popup object.
# Notice we're referencing the column names with the '$', same as we would with a non-spatial dataframe.
economyPopup <- paste0("<strong>Country: </strong>", 
                world$name, 
                "<br><strong>Market Stage: </strong>", 
                 world$economy)
```


Now we can use leaflet to build a world map that is shaded by GDP and displays a popup. Note the 'layerId = ~name' snippet below - it creates a layer of country names. We will change that later in an important way.

```{r}
# Build our leaflet map object.

leaf_world_economy <- leaflet(world) %>%
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng =  20, lat =  15, zoom = 2) %>%
      addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = .7, color =
      ~gdpPal(gdp_md_est), layerId = ~name, popup = economyPopup)

# Display that object below.

leaf_world_economy
```


The map looks good, but it sure would be nice if we could add the ETF ticker symbols and year-to-date data to the world spatial dataframe object - and we can! Our ‘name’ column in the ETF dataframe uses the same country naming convention as the ‘name’ column of the map, and those columns are both called ‘name’. Thus, we can use the merge() function from the ‘sp’ package to add the ETF dataframe to the spatial dataframe. This is similar to a ‘join’ using dplyr.

The correspondence of country names wasn’t just luck - I had the benefit of having worked with this shapefile in the past, and made sure the country names matched up, and now you have the benefit of having worked with this shapefile. For any future project that incorporates a map like this, give some forethought to how data might need to be merged with the shapefile. The shapefile and the new data need a way to be matched. Country names usually work well.

After the merging, the ticker symbols and year-to-date number columns will be added for each country that has a match in the ‘name’ column. For those with no match, the ‘ticker’ and ‘ytd’ columns will be filled with NA. 

```{r}
library(sp)
# Once we run this line of code, our ticker symbols and ytd numbers will be added
# to the spatial dataframe.

world_etf <- merge(world, etf_ticker_country, by = "name")
```

Now that the ytd data is added, let’s shade the different countries according to the year-to-date performance of the country EFT, instead of by GDP as we did before. A nice side benefit of this new shading scheme: if a country has no ETF, it will remain an unattractive grey.

```{r}
# Create a palette with different shades of blue for different
# year-to-date performances. Previously, we shaded by 'world$gdp_md_est', now
# we'll shade by 'world_etf$ytd'.

ytdPal <- colorQuantile("Blues", world_etf$ytd, n = 20)
```


The new shading is nice, but let’s also have the popup display the exact year-to-date performance percentage for any detail-oriented users.

```{r}
# Create a popup that displays the year-to-date performance.

ytdPopup <- paste0("<strong>Country: </strong>", 
                world_etf$name,
                "<br><strong> Year-to-date: </strong>", 
                world_etf$ytd, "%")
```
 
Now we’ll build another map that uses the year-to-date color scheme and popup, but we will make one more massively important change: we will change ‘layerId = ~name’ to ‘layerId = ~ticker’ to create a map layer of tickers.

Why is this massively important? When we eventually create a Shiny app, we want to pass ticker symbols to ‘getSymbols’ based on a user click. The ‘layerId’ is how we’ll do that:  when a user clicks on a country, we capture the ‘layerId’, which is a ticker name that we can pass to ‘getSymbols.’ But that is getting ahead of ourselves. For now, here is the new map:

```{r}

leaf_world_etf <- leaflet(world_etf) %>%
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng =  20, lat =  15, zoom = 2) %>%
      addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = .7,
                  
      # The next line of code is really important for creating the map we want to use later.       
      
      color =~ytdPal(ytd), layerId = ~ticker, popup = ytdPopup)

leaf_world_etf
```


This map is going to play a crucial role in our Shiny app and the last step is to save the shapefile, that we confirmed works well with leaflet, for use in our flexdashboard. 


```{r}
# Save the shapefile but NOT the ETF price data to a .RDat file.
# We'll load this file and use it in our Shiny app.

save(world_etf, file = 'etfData.RDat')
```

On to the Shiny app (which, in real life, would require its own separate Rmd file - the remaining code chunks are not meant to be run in this Notebook and all include eval = FALSE for that reason; they must be run in the separate Rmd file with runtime: shiny). 

We will use the flexdashboards package to create this app. If you are accustomed to building Shiny apps in app.r files, this will be a nice introduction to flexdashboards - no more UI and server logic, as we'll see! 

First, a quick look at the YAML header. 

```{r, eval = FALSE}
title: "Global ETF Map"
runtime: shiny
output:
 flexdashboard::flex_dashboard:
  source_code: embed
```

Now, on to the first code chunk, where we load the ‘etfData.RDat’ file, the same file where we saved our shapefile of the world. After loading that file, we use the exact same code from the Notebook to construct our leaflet map.

```{r, eval = FALSE}
load('etfData.RDat')

# The next lines are taken straight from the Notebook. We just need
# to rebuild that map.

ytdPal <- colorQuantile("Blues", world_etf$ytd, n = 20)

ytdPopup <- paste0("<strong>Country: </strong>", 
                world_etf$name,
                "<br><strong> Year-to-date: </strong>", 
                world_etf$ytd, "%")


leaf_world_etf <- leaflet(world_etf) %>%
  addProviderTiles("CartoDB.Positron") %>% 
  setView(lng =  20, lat =  15, zoom = 2) %>%
      addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = .7,
                  
      # The next line of code is really important!       
      
      color =~ytdPal(ytd), layerId = ~ticker, popup = ytdPopup)

```

Next, we want to display that map in the first row of our flexdashboard. We will use ‘renderLeaflet’. Nothing fancy here.

```{r, eval = FALSE}
leafletOutput("map1")

output$map1 <- renderLeaflet({
    leaf_world_etf
})
```

Now, we want to add the ability for the user to click a country and see the price history of that country’s ETF. This will require the following steps: (1) click a country, (2) pass the ETF ticker symbol to getSymbols, (3) import the price data, and (4) display the timeseries using dygraphs. All of this is going to be easy for us because of an important decision that we made when building our leaflet map object way back in the Notebook.

Recall that we set layerID = ticker when we added our polygons. This is part of the magic of the shapefile. It allows us to capture the ticker symbol associated with a country when a user clicks. In the code chunk below, we use an observeEvent function to capture the ID of whatever shape a user clicks, and we have set that ID to be the ticker.

```{r, eval = FALSE}
clickedCountry <- eventReactive(input$map1_shape_click, {
     return(input$map1_shape_click$id)
     })
```

Now, we have the ticker in the ‘clickedCountry’ reactive and can pass it to getSymbols. One important note: if a user clicks on a country with no ETF, that reactive object will be ‘NA’ or null. The first line in the code chunk below is to validate the entry and, if it is null, let the user know that a country with no ETF has been clicked.

```{r, eval = FALSE}
validate(need(clickedCountry(), "There is no iShares ETF for that country"))

etf <- getSymbols(as.character(clickedCountry()), auto.assign = FALSE)
```

To graph the resulting xts object, which we have named ‘etf’, we pass it to dygraph.

```{r, eval=FALSE}
dygraph(Ad(etf), main = clickedCountry())
```

That’s the final code chunk and we're done! But keep in mind: this interface need not be limited to country ETFs. We could wire this up to display currency prices, interest rates, labor force participation rates, or, say, the 10 best-performing stocks in a country. Once we have our template to use a map as an interface to data, the world is our playground.